A Practical Optional Type System for Clojure

### 2012-10-17 dissertation/design-choices.tex
Lots of stuff
Implementation work

Typed Clojure is designed to be of practical use to Clojure programmers.
Many of the design goals are similar to Tobin-Hochstadt's \cite{Tob10}
for Typed Racket.

@@

In Java, \lstinline|null| is a subtype to all reference
types, but \lstinline|null| is not expressible as a static type. In other words,
when a Java type claims a reference type, you should also assume it can also be of 
type \lstinline|null|.

Scala provides the Option type for safe \lstinline|null| elimination, forcing pattern matching
where \lstinline|null| might occur.

@@

>>>>

Typed Clojure is designed to be of practical use to Clojure programmers.
Many of the design goals are similar to Tobin-Hochstadt's 
for Typed Racket \cite{Tob10}. Some of the main differences are described in this chapter.

@@

In Java, \lstinline|null| is included in all reference
types, but \lstinline|null| is not expressible as a static type. In other words,
when a Java type claims a reference type, you should also assume it can also be of 
type \lstinline|null|.

Scala provides the Option type for safe \lstinline|null| elimination, forcing pattern matching
where \lstinline|null| might occur, and

@@

As a key idiomatic feature of Clojure, we treat heterogeneous maps more formally.
This also serves as a prototype for formalising other aspects.
Here we give a formal syntax, operational semantics and type system using inference
rules, as is standard in the programming language community (see, eg., Pierce\cite{Pie02}).

### 2012-10-17 dissertation/implementation.tex
Lots of stuff
Implementation work

This chapter discusses the implementation of the Typed Clojure prototype type 
system\footnote{We refer to this implementation as \emph{Typed Clojure} for the remainder of this chapter.}.
concentrating on significant challenges that were identified and overcome.
Many aspects of Typed Clojure's design follows the implementation of Typed Racket

@@

Typed Clojure and Typed Racket differ significantly in how they interact with
the Clojure and Racket respectively to perform type checking.
Typed Racket is implemented as a language on the
Racket platform, which provides highly sophisticated and extensible macro facilities.
This allows Typed Racket to be entirely implemented with macros.
Instead, Typed Clojure is implemented as a library that utilises abstract syntax trees (AST)
generated by analyze, a library developed for this project.

@@

\section{Occurrence typing}

Typed Clojure's implementation of occurrence typing is ported and extended from Typed Racket.
Occurrence typing plays several roles in Typed Clojure.
First, occurrence typing is used to update the type environment at every conditional branch.
Second, it is used to calculate whether branches are reachable: a branch is not reachable
if updating the type environment results in a variable of the bottom type.
Third, paths are used extensively in Typed Clojure. Typed Racket uses \emph{car} and \emph{cdr} paths
to track which component of a cons cell to update. Typed Clojure adds paths for traversing heterogeneous maps,
inferring length information, and \lstinline|first| and \lstinline|rest| paths for sequences.

@@

@@

A significant portion of Typed Clojure is theoretically platform independent but there are
challenges to targeting new dialects, including incompatible host interoperability and
non-standardised abstract syntax trees.
The first issue is predictable due to a core philosophy of Clojure dialects: 
host interoperability is non-portable\footnote{See the complete Clojure rationale: http://clojure.org/rationale}.
Each dialect of Clojure has a unique host interopability story and Typed Clojure should cater for them
separately.
The second issue is potentially resolvable.

@@

>>>>

This chapter discusses the implementation of the Typed Clojure prototype type 
system\footnote{We refer to this implementation as \emph{Typed Clojure} for the remainder of this chapter.}
in some detail concentrating on significant challenges that were identified and overcome.
Many aspects of Typed Clojure's design follows the implementation of Typed Racket,
which is reflected in this chapter.

@@

Typed Clojure and Typed Racket differ significantly in how they perform type checking.
Typed Racket is implemented as a language on the
Racket platform, which provides highly sophisticated and extensible macro facilities.
Interestingly, this allows Typed Racket to be entirely implemented with macros.
Instead, Typed Clojure is implemented as a library that utilises abstract syntax trees (AST)
generated by analyze\footnote{https://github.com/frenchy64/analyze}, a library I developed for this project.
This strategy follows common practice for Clojure projects, which favours providing modular libraries 
over modifying the language.

This section goes into details on the implementation of Typed Clojure.
First, a high-level overview is given on the type checking procedure.
Then the interfaces to particular high-level functions are discussed.

\subsection{General Overview}

@@

\subsection{Bidirectional Checking}

The interface to the bidirectional checking algorithm is \lstinline|typed.core/check|,
which takes an expression, represented as an AST generated from \emph{analyze}, and an optional expected type for
the given expression. If the expected type is present, the bidirectional algorithm \emph{checks}
that the expected type matches the actual type of the expression.
If the expected type is omitted, a type is instead \emph{synthesised} for the expression.
The algorithm is based on Pierce and Turner's  Local Type Inference\cite{PT00}
and the implementation is similar in shape to Typed Racket's\cite{TF08}, in that
one function with an extra ``expected type'' argument is preferred over two complementary
functions, one for checking and one for synthesis.

The main difference between Typed Clojure's and Typed Racket's bidirectional checking
algorithm is the representation for expressions. Typed Racket relies on pre-existing
Racket features like syntax objects for expression representation. Clojure instead
leans towards abstract syntax tree representation, despite its Lisp heritages.
In terms of the bidirectional check, the difference is mostly cosmetic.

% go into more detail
% - update env
% - what is an environment? Bindings + propositions
% - LOC
% - link to github
% - how occurrence typing works
%   - update env
%   - how to calculate reachability

\subsection{Occurrence typing}

Typed Clojure's implementation of occurrence typing is ported and extended from Typed Racket.
Occurrence typing plays several roles in Typed Clojure.
First, occurrence typing is used to update the type environment at every conditional branch.
Second, it is used to calculate whether branches are reachable.
Third, paths are used extensively in Typed Clojure. 
The implementation of these features are discussed in this section.

The basic idea of occurrence typing involves keeping a separate environment of \emph{propositions}
that relate bindings to types. These propositions are then used to update the type environment.
In Typed Clojure there are several types of propositions, referred to as \emph{filters}.
They are based on the theory by Tobin-Hochstadt and Felleisen\cite{TF10}, and are ported directly
from Typed Racket.

\begin{itemize}
  \item TopFilter and BotFilter represent the trivially true and trivially false propositions respectively.
  \item TypeFilter and NotTypeFilter represent a positive or negative association of a binding name
        to a type. The syntax \lstinline|(is t name)|
  \item AndFilter and OrFilter represent logical conjunction and disjunction of propositions.
\end{itemize}

Propositions can optionally carry \emph{path} information
represented by a sequence of \emph{path elements}, which signify which part
of the binding's type to update. For example, Typed Racket uses \emph{car} and \emph{cdr} path elements
to track which component of a cons type to update.
Paths are particularly useful in Typed Clojure. There are path elements for traversing heterogeneous map types (KeyPE),
inferring length information (CountPE), and \lstinline|first| and \lstinline|rest| paths for sequences.
These additions do not appear to introduce any significant new complexities related to paths.

\section{Polymorphic Type Inference}

The polymorphic type inference algorithm is directly ported from Typed Racket with slight extensions
for bounded variables,
and is reminiscent to Pierce and Turner's Local Type Inference\cite{PT00}.
A common entry point for inferring type variables to polymorphic function invocations
\lstinline|typed.core/infer|.

\lstinline|infer| is invoked like \lstinline|(infer X Y S T R expected)|, where

\begin{itemize}
  \item \lstinline|X| is a map from type variable names
        to their bounds (representing the type variables in scope),
  \item \lstinline|Y| is a map from type variable names
        to their bounds (representing the \emph{dotted} type variables in scope),
  \item \lstinline|S| and \lstinline|T| are sequences of types of equal length
  \item \lstinline|R| is a type,
  \item \lstinline|expected| is the expected type for R, or the value \lstinline|nil|,
\end{itemize}

and returns a \emph{substitution} that satisfies the following conditions:

\begin{itemize}
  \item Pairwise, each \lstinline|S| is a below \lstinline|T|,
  \item \lstinline|R| is below \lstinline|expected|, if \lstinline|expected| is provided.
\end{itemize}

A substitution maps type variables to an upper and lower bound. 
It is valid to replace all occurrences of the type variables named in the substitution 
with a type between their associated bounds.
For example, usually the substitution is applied to \lstinline|R| by the caller of \lstinline|infer|.
Whether to substitute the upper or lower bound is determined by the variance
of the type variable in relation to the type being substituted. 
The specific algorithm is described in Local Type Inference.

\section{F-Bounded Polymorphism}

A feature not present in Typed Racket is bounded polymorphism.
Several changes were needed to support bounded polymorphism. In every position
where a set of type variables was required, it was replaced by a map
of type variables to bounds.

Bounds consist of an upper and lower type bound, or a \emph{kind bound}.
Kind bounds are experimental following the inclusion of user definable type constructors
(motivated in section \ref{sec:monads}).
They are a stub for a more comprehensive treatment of higher-kinded operators
such as that described by Moors, Piessens, and Odersky for Scala\cite{MPO08}.
At present, a type variable can only be instantiated to a type between
its upper and lower bounds, or, if a kind bound is defined instead, 
to a kind below the kind bound.

F-bounded polymorphism allows type variables to refer to themselves in their bounds.

@@

It also required changes to the polymorphic type inference algorithm.
The constraint-generation portion

@@

A significant portion of Typed Clojure is theoretically platform independent but there are
challenges to targeting new dialects, including incompatible host interoperability and
non-standardised abstract syntax trees.
The first issue is predictable due to a core philosophy of Clojure dialects: 
host interoperability is non-portable\footnote{See the complete Clojure rationale: http://clojure.org/rationale}.
Each dialect of Clojure has a unique host interoperability story and Typed Clojure should cater for them
separately.
The second issue is potentially resolvable either by enforcing a standard representation for 
abstract syntax trees across Clojure implementations, or developing a library that provided
a common interface to abstract syntax trees for each Clojure implementation.

@@

\section{Filtering on Negative Propositions}
\label{sec:filterneg}

### 2012-10-17 dissertation/intro.tex
Implementation work

In the last decade it has become increasingly common to enhance
dynamically typed languages with static type systems. This idea not new,
but recent attempts are noteworthy for their successful use of bidirectional checking.
Instead of always attempting to infer types the algorithm relies on programmer annotations
appearing in some natural places, such as giving the type of each top-level function.

The Clojure programming language is a dynamically typed language running on the Java Virtual Machine. 
It emphasises functional programming with immutable data structures
and provides direct interoperability with languages on the Java Virtual Machine.
Clojure also is also a family of languages with similar semantics, referred to as Clojure dialects.
There are several Clojure dialects, notably the aforementioned Clojure for the Java Virtual Machine (JVM),
ClojureCLR which targets the Common Language Runtime, and ClojureScript which compiles to Javascript.
Where it is clear to do so, we refer to Clojure the programming language as \emph{Clojure}, 
and the family of Clojure-like languages as \emph{Clojure dialects}.

Clojure dialects are widely used for several reasons.
Performance is a key feature, for example the JVM implementation of Clojure
offers ways to access Java-like speed for certain operations.

Recently languages have been created to or been modified to support aspects
of both static and dynamic typing.
Dart is fundamentally dynamically typed but offers a form of optional
static typing. Haskell recently supports the ability to delay static type errors 
to runtime, removing the need for the program to type check before running.
Scala can emulate some features of dynamic languages with Dynamic Scala.

@@

% What are types? No real agreement .. but here is my definition..
% Every programmer thinks they know what they are. More than one concept
% in the literature.
% Static vs. dynamic
% intrinsic vs. extrinsic

% is it even a type system?
% type refinements vs ordinary types
% extrinsic vs intrinsic type systems

% refinements check invariants of programs
% while ordinary static types check whether programs
% are basically meaningful - Rowan + Frank P
% Page 1 Rowan's PhD
\begin{verbatim}Following the terminology of [Pfenning, Reynolds ..]\end{verbatim}
An ordinary static type system is used to check whether programs are basically
meaningful. Such type systems are \emph{intrinsic}. A language with an intrinsic
static type system has run-time semantics that depends on the types of associated
variables and expressions during type checking.
For example, C, Java and ML have intrinsic types.
This means such programs must pass the type checker before being run.

A static type system is \emph{extrinsic} when runtime semantics
does not depend on the type system. In other words, passing a static type checker
is not essential to running programs. A dynamically typed language can be viewed
as having a trivial static type system that supports exact one type,
\begin{verbatim} .. a view strongly advocated by Harper[..]
and common in the literature on static types (see, eg. Pierce[..])
\end{verbatim}

A type checker ensures invariants hold complains if code
does not conform to its documented type. Powerful type systems like
Haskell's and Scala's are useful when reasoning about higher-order code.

When a static type checker is not available, which describes the situation
for most dynamic languages, other techniques are used for checking
type invariants. Runtime contracts and unit testing are a popular alternatives.
Clojure adopts these approaches, providing easy syntax for defining pre-
and post-conditions \footnote{Inspired by Eiffel} and a library for writing unit tests.
\begin{verbatim} but.. (why a static type system is still desirable\end{verbatim}

% clojure is dynamically typed
% currently few tools to statically reason about complex higher-order code
% conduit based on streams, uses continuations
% core.logic based on monads
% Static types are useful for reasoning for such things
% Strong statically typed languages like Haskell use monads regularly

% strength of Clojure is concision
% typically, concise static languages are built concurrently with their type system
% Eg. Haskell, ML flavoured languages
% Counterpoint: Typed Racket preserves racket idioms, adds static typing

% Clojure community reactions
% GSOC
% Conj talk
% 
% Motivation: Conduit

@@

% Hello world
This section introduces Typed Clojure with example code. An attempt is
made to introduce some Clojure syntax and semantics to those unfamiliar or needing a refresher.
A basic knowledge of Lisp syntax is handy, but a brief tutorial is given
for newcomers.

@@

The core of understanding Lisp syntax can be summarised by these points

\begin{itemize}
  \item Invocations are always wrapped in a pair of balanced parenthesis 
  \item Parenthesis start to the left of the operator
\end{itemize}

@@

@@

\begin{lstlisting}[caption=Hello world]
(ns typed.test.hello-world
  (:require [typed.core :refer [check-ns]]]))

@@

Here \lstinline|clojure.core/ns|
\footnote{All vars from the \lstinline|clojure.core| namespace are referred implicitly in new namespaces (equivalent to \lstinline|(:require [clojure.core :refer :all])|).}
declares the current namespace to be 
\lstinline|typed.test.hello-world|
and requires the namespace \lstinline|typed.core| to be loaded as a dependency,
referring \lstinline|typed.core/check-ns| to be referenced as \lstinline|check-ns|. Other than this
dependency, this is identical to the untyped \emph{Hello world}.
By running \lstinline|check-ns| we can verify that this namespace type checks.

@@

\begin{lstlisting}
(ns typed.test.collatz
  (:require [typed.core :refer [check-ns ann]]))

@@

\begin{lstlisting}
(ns typed.test.deftype
  (:require [typed.core 
             :refer [check-ns ann-datatype
                     tc-ignore ann-protocol AnyInteger]]))

@@

\begin{lstlisting}
...
(ann to-set 
     (All [x]
       [(U nil (Seqable x)) -> (clojure.lang.PersistentHashSet x)]))
(defn to-set [a]
  (set a))
...
\end{lstlisting}

@@

\begin{lstlisting}
...
(ann config
     (HMap {:file String,
            :ns Symbol}))
(def config
  {:file "clojure/core.clj",
   :ns 'clojure.core})
...
\end{lstlisting}

@@

\begin{lstlisting}
(ns typed.test.poly
  (:require [typed.core :refer [ann AnyInteger check-ns cf]])
  (:import [clojure.lang Seqable]))

@@

\begin{lstlisting}
(ann num-vec2 
     [(U nil Number) (U nil Number) -> (Vector* Number Number)])
(defn num-vec2 [a b]
  [(if a a 0) 
   (if b b 0)])
\end{lstlisting}

@@

\begin{lstlisting}
(ns typed.test.interop
  (:import (java.io File))
  (:require [typed.core :refer [ann non-nil-return check-ns]]))

>>>>

In the last decade it has become increasingly common to enhance
dynamically typed languages with static type systems. 
This idea not new, but recent attempts are noteworthy for their successful use of bidirectional checking\footnote{These 
slides by Felleisen, a practitioner in this area for over 20 years, present a history of optional type checking
for Lisp/Scheme that spans around 40 years\cite{Fell09}}.
Instead of always attempting to infer types the algorithm relies on programmer annotations
appearing in some natural places, such as giving the type of each top-level function.

The Clojure programming language is a dynamically typed dialect of Lisp invented
by Hickey, designed to run on popular platforms\footnote{http://clojure.org/}.
It emphasises functional programming with immutable data structures
and provides direct interoperability with its host platform.
Notable implementations of Clojure exist for the Java Virtual Machine (JVM),
the Common Language Runtime, and for Javascript virtual machines.
At the current time, Clojure on the JVM is the most mature implementation,
and therefore this project focuses on the JVM implementation.

Clojure has attracted wide-spread users in part by concentrating on pragmatism.
Performance is a key feature, for example the JVM implementation of Clojure
offers ways to access Java-like speed for certain operations.
Also, Clojure's excellent host interoperability offers Clojure programmers
access to existing libraries for their platform, such as the extensive Java library ecosystem.
By coupling pragmatic necessities with elegant features like Lisp, functional programming,
and immutability by default, Clojure is a compelling general purpose programming language.

Recently languages have been created to or been modified to support aspects
of both static and dynamic typing.
Dart\footnote{http://www.dartlang.org/} is dynamically typed but offers optional type annotations
that do not affect runtime semantics. 
Typescript\footnote{http://www.typescriptlang.org/} adds an optional type system to Javascript,
a well-known dynamic language.

When a static type checker is not available, which describes the situation
for most dynamic languages, other techniques are used for checking
type invariants. For example, ``design by contract'' is often used,
introduced by Meyer for the Eiffel language\cite{Mey92},
in which the programmer defines contracts that are enforced at runtime.
Unit testing is also a popular verification technique in dynamic languages.
Clojure adopts these approaches, providing easy syntax for defining 
Eiffel-inspired pre- and post-conditions and a library for writing unit tests.

Static type systems, however, are still desirable for certain situations.
Powerful type systems like ML's\cite{Mil97} and Haskell's\cite{Mar10} have proved particularly useful
for writing in complicated programming styles. For example,
Haskell's advanced static type system helps the programmer write correct monadic code,
especially in more complicated situations like combining monads via monad transformers.

\subsection{Why implement an optional type system for Clojure?}

The initial motivation for implementing an optional type system
for Clojure occurred during an anecdotal account of the struggles
of a Clojure programmer. In a heroic effort, he managed to 
implement a library for conduits (or \emph{arrows}) in Clojure, 
which probably surpass monads in complexity, and is usually reserved
for languages with advanced type systems like Haskell.

He highlighted a strong desire for a type system for several reasons.
Firstly, to verify the correctness
of the library (before noting that unit tests and runtime contracts
were insufficient for this task).
Secondly, to aid him while writing the library.

Since then, the Clojure community has shown interest in this work.
I developed this project as a Google Summer of Code 2012 project,
after it was selected by the Clojure community.
I have also been invited to speak on this project at the Clojure Conj
2012 conference, a major annual Clojure conference.

@@

% refinements check invariants of programs
% while ordinary static types check whether programs
% are basically meaningful - Rowan + Frank P
% Page 1 Rowan's PhD

There are many concepts associated with \emph{types} and \emph{type systems} in both the
literature and informal discourse.
A programmer who uses dynamically-typed languages may have a drastically different notion
of what a type is than, say, a programmer preferring languages with advanced static type systems.
There is some debate as to whether optional static type systems like Typed Clojure
can even be called a type system. We choose to
follow the terminology of Pfenning\cite{Pfe08} and Reynolds\cite{Rey01},
where such optional type systems are \emph{extrinsic} type systems, and more
traditional type systems are \emph{intrinsic} type systems.

An ordinary static type system is used to check whether programs are basically
meaningful. Pfenning and Reynolds call these type systems \emph{intrinsic}. A language with an intrinsic
static type system has run-time semantics that depends on the types of associated
variables and expressions during type checking.
For example, C, Java, and ML have intrinsic types.
This means programs written in these languages must pass the type checker before being run.

A static type system is \emph{extrinsic} when runtime semantics
does not depend on the type system. In other words, passing a static type checker
is not essential to running programs. A dynamically typed language can be viewed
as having a trivial static type system that supports exact one type,
a view strongly advocated by Harper\cite{Har12}
and common in the literature on static types (for example, Pierce\cite{Pie02}).

@@

% Hello world
This section introduces Typed Clojure with example code. 
Typed Clojure is developed for the JVM implementation of Clojure, therefore
the rest of this chapter uses that implementation.
An attempt is
made to introduce some Clojure syntax and semantics to those unfamiliar or needing a refresher.
A basic knowledge of Lisp syntax is handy, but a brief tutorial is given
for newcomers.

@@

The core of understanding Lisp syntax when coming from a popular language
like Java or Javascript can be summarised by these points.

\begin{itemize}
  \item Operators are always in prefix position.
  \item Invocations are always wrapped in a pair of balanced parenthesis.
  \item Parenthesis start to the left of the operator.
\end{itemize}

@@

Clojure also adds other syntax:

\begin{itemize}
  \item Prefixing \lstinline|:| to a symbol defines a \emph{keyword}, often used for map keys. eg. \lstinline|:my-keyword|.
  \item Square brackets delimit vector literals. eg. \lstinline|[1 2]| is a 2 place vector.
  \item Curly brackets define map literals. eg. \lstinline|{:a 1 :b 2}| is a map from 
        \lstinline|:a| to \lstinline|1| and \lstinline|:b| to \lstinline|2|.
  \item Commas are always optional, and treated as whitespace.
\end{itemize}

@@

\begin{lstlisting}[caption=Typed Hello world, label=lst:helloworld]
(ns typed.test.hello-world
  (:require [typed.core :refer [check-ns]]]))

@@

At this point, it is worth understanding Clojure's namespacing feature.
Clojure code is always executed in a \emph{namespace}, and each file of Clojure code should 
have a \lstinline|ns| declaration with the namespace name and its dependencies,
which switches the current namespace and executes the given dependency commands.
There is one special namespace, \lstinline|clojure.core| which is
loaded with every namespace, implicitly ``referring'' all its vars in the namespace.
For example, \lstinline|ns| refers to the var \lstinline|clojure.core/ns|,
similarly \lstinline|println| refers to \lstinline|clojure.core/println|.

The example in listing \ref{lst:helloworld} declares a dependency to 
\lstinline|typed.core|, Typed Clojure's main namespace. It also refers the var \lstinline|typed.core/check-ns|
into scope\footnote{\lstinline|check-ns| is the top level function for type checking a namespace}.
Other than this dependency, this is identical to the untyped \emph{Hello world}.

@@

\begin{lstlisting}[caption=Annotating vars in Typed Clojure]
(ns typed.test.collatz
  (:require [typed.core :refer [check-ns ann]]))

@@

\begin{lstlisting}[caption=Annotating protocols and datatypes in Typed Clojure]
(ns typed.test.deftype
  (:require [typed.core 
             :refer [check-ns ann-datatype
                     tc-ignore ann-protocol AnyInteger]]))

@@

\begin{lstlisting}[caption=Polymorphism in Typed Clojure]
...
(ann to-set 
     (All [x]
       [(U nil (Seqable x)) -> (clojure.lang.PersistentHashSet x)]))
(defn to-set [a]
  (set a))
...
\end{lstlisting}

@@

\begin{lstlisting}[caption=Heterogeneous map types in Typed Clojure]
...
(ann config
     (HMap {:file String,
            :ns Symbol}))
(def config
  {:file "clojure/core.clj",
   :ns 'clojure.core})
...
\end{lstlisting}

@@

\begin{lstlisting}[caption=Annotating multi-arity functions in Typed Clojure]
(ns typed.test.poly
  (:require [typed.core :refer [ann AnyInteger check-ns cf]])
  (:import [clojure.lang Seqable]))

@@

\begin{lstlisting}[caption=Example of occurrence typing in Typed Clojure]
(ann num-vec2 
     [(U nil Number) (U nil Number) -> (Vector* Number Number)])
(defn num-vec2 [a b]
  [(if a a 0) 
   (if b b 0)])
\end{lstlisting}

@@

\begin{lstlisting}[caption=Java interoperability with Typed Clojure]
(ns typed.test.interop
  (:import (java.io File))
  (:require [typed.core :refer [ann non-nil-return check-ns]]))

### 2012-10-17 dissertation/lit-review.tex
Implementation work

Typed Clojure is related to several existing systems for both typed and untyped languages.
The first section of this chapter gives an overview of type systems designed for untyped languages.
The second section explores how existing type systems statically type check multimethods
and their relevance to Clojure's multimethods.

@@

\section{Intersection Types}

Intersection types are an interesting type construct relevant to capturing the complicated
types manipulated by dynamic languages.

@@

% CIDRE
SML CIDRE is a refinement type checker for Standard ML by Davies \cite{Dav05}
that features a form of intersection types. It is used for function overloading
and checking invariants involving Refinement types.
\begin{verbatim}
Freeman & Pfenning, Refinement types
\end{verbatim}

Typed Racket, developed by Tobin-Hochstadt and others \cite{Tob10}, uses \emph{ordered intersection types} 
to offer a kind of function overloading.
\begin{verbatim}PADL '12 paper, Typing the Numeric Tower\end{verbatim}

Clojure provides two dispatching mechanisms: multimethods, a mechanism to dispatch on the
results of an arbitrary function; and protocols, a mechanism for single dispatch, a common
strategy that dispatches on the type of the first provided argument.
Millstien and Chambers explore statically typing multiple dispatched multimethods in Dubious \cite{MS02}.
Dubious' multimethods are less powerful than Clojure's multimethods but more powerful than protocols,
and Clojure's dispatch functions can be extended at runtime, unlike Dubious' multimethods.

Clojure and the statically typed language Scala \cite{OCD+} 
have some interesting similarities. The redesign
of the Scala collection hierarchy\cite{OS10}
bears similarity to Clojure's encompassing ``seq''
model, revealing an opportunity to test potential typed extensions to the Clojure 
class hierarchy.

Scala and Clojure are both languages with a focus on host interoperability, specifically
to the Java Virtual Machine. Their differing treatment of the Java Virtual Machine's \lstinline|null| is
relevant, particularly considering the potential of using occurrence typing \cite{TF10}
to prevent erroneous usage of \lstinline|null|.

@@

In typed languages like Haskell, functions are as simple as possible, taking a single argument.
A function with multiple arguments is represented by chaining several single-argument function
together. This style is characterised by direct syntactic function currying, where applying
a function to less than its maximum number of arguments results in another function
that takes the remaining arguments.

>>>>

%Typed Clojure is related to several existing systems for both typed and untyped languages.

@@

\section{Intersection and Union Types}

Intersection and union types are interesting type constructs relevant to capturing the complicated
types common in dynamic languages.
An intersection of types \lstinline|a| and \lstinline|b|, written \lstinline|(I a b)| in Typed Clojure,
can be used safely in both positions expecting type \lstinline|a|,
and positions expecting type \lstinline|b|.
Intuitively, \lstinline|(I a b)| is both of type \lstinline|a| \emph{and} \lstinline|b|.
A union of types \lstinline|a| and \lstinline|b| is \emph{either} of type \lstinline|a| or
\lstinline|b|, written \lstinline|(U a b)| in Typed Clojure,
and can be used safely in positions expecting a type that is \emph{both} type \lstinline|a|
and \lstinline|b|.
For example, \lstinline|(U Number Boolean)| cannot be used in positions expecting \lstinline|Number|.

Several interesting projects have used intersection or union types.

@@

SML CIDRE is a refinement type checker for Standard ML by Davies \cite{Dav05}
that features a form of intersection types. It is used for function overloading
and checking invariants involving refinement types.
\begin{verbatim}
Freeman & Pfenning, Refinement types
\end{verbatim}

St-Amour, Tobin-Hochstadt, Flatt, and Felleisen
describe the \emph{ordered intersection types} used in Typed Racket\cite{St12}
that provide a kind of function overloading.
Typed Clojure takes a similar approach
\begin{verbatim}
described in chapter
\end{verbatim}
Typed Racket also uses union types

\section{Java Interoperability}

Scala and Clojure are both languages with a focus on host interoperability, specifically
to the Java Virtual Machine. Their differing treatment of the Java Virtual Machine's \lstinline|null| is
relevant, particularly considering the potential of using occurrence typing \cite{TF10}
to prevent erroneous usage of \lstinline|null|.
\begin{verbatim}
eg.
\end{verbatim}

@@

In typed languages like Haskell, functions are as simple as possible, taking a single argument.
A function with multiple arguments is represented by chaining several single-argument function
together, or by using lists, or using tuples. This style is characterised by direct syntactic function currying, where applying
a function to less than its maximum number of arguments results in another function
that takes the remaining arguments.

### 2012-10-17 dissertation/paper.tex
Implementation work

\usepackage{listings}
\lstset{ %
  language=Lisp,                % choose the language of the code
  columns=fixed,basewidth=.5em,
  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
  %numbers=left,                   % where to put the line-numbers
  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
  %numbersep=5pt,                  % how far the line-numbers are from the code
  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  %showspaces=false,               % show spaces adding particular underscores
  showstringspaces=false,         % underline spaces within strings
  %showtabs=false,                 % show tabs within strings adding particular underscores
  %frame=single,           % adds a frame around the code
  %tabsize=2,          % sets default tabsize to 2 spaces
  captionpos=t,           % sets the caption-position to bottom
  breaklines=true,        % sets automatic line breaking
  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
}

@@

This paper describes my work on designing an optional static type system for Clojure based on
the lessons learnt from several projects, primarily Typed Racket.
A prototype type system is developed for Clojure running on the Java Virtual Machine.
This prototype is sufficient to allow experiments with existing Clojure code that is sufficiently complicated
that type checking increases confidence that the code is correct.
For example, a significant portion of algo.monads, a Clojure Contrib library for monadic programming, 
is able to be type checked. 
On the positive side, most monad, monad transformer, and monadic function definitions can be type checked,
usually by adding type annotations in natural places like function definitions.
Negatively, simple monad comprehensions can be type checked, but require awkward type annotations.

There is significant future work to fully type check all Clojure features and idioms.
For example, multi-method definitions and functions accepting an even number of variable arguments
are troublesome. 
Also, there are desirable features from the Typed Racket project that are missing, such
as automatic runtime contract generation and a sophisticated blame system, 
both which significantly improve error messages when mixing typed and untyped code.
I conclude that it appears to be both practical and useful to design and implement an optional static type system for the
Clojure programming language.

>>>>

\usepackage{listings}
\lstset{ %
  language=Lisp,                % choose the language of the code
  columns=fixed,basewidth=.5em,
  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
  %numbers=left,                   % where to put the line-numbers
  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
  %numbersep=5pt,                  % how far the line-numbers are from the code
  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  %showspaces=false,               % show spaces adding particular underscores
  showstringspaces=false,         % underline spaces within strings
  %showtabs=false,                 % show tabs within strings adding particular underscores
  frame=single,           % adds a frame around the code
  %tabsize=2,          % sets default tabsize to 2 spaces
  captionpos=t,           % sets the caption-position to bottom
  breaklines=true,        % sets automatic line breaking
  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
}

@@

This paper describes my work on designing an optional static type system for Clojure based on
the lessons learnt from several projects, primarily Typed Racket.
A prototype type system is developed for Clojure running on the Java Virtual Machine.
Several experiments are conducted using this prototype, particularly involving
sufficiently complicated existing Clojure code where type checking increases confidence that the code correct.
For example, a significant portion of algo.monads, a Clojure Contrib library for monadic programming, 
is able to be type checked. 
On the positive side, most monad, monad transformer, and monadic function definitions can be type checked,
usually by adding type annotations in natural places like function definitions.
Negatively, simple monad comprehensions can be type checked, but require awkward type annotations.

There is significant future work to fully type check all Clojure features and idioms.
For example, multimethod definitions and functions accepting an even number of variable arguments
are troublesome. 
Also, there are desirable features from the Typed Racket project that are missing, such
as automatic runtime contract generation and a sophisticated blame system, 
both which are designed to improve error messages when mixing typed and untyped code in similar systems.
I conclude that it appears to be both practical and useful to design and implement an optional static type system for the
Clojure programming language.
